---
title: "Exam 2"
author: "Zack Niedzwiecki"
format: pdf
editor: visual
---

## Results

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
library(ggplot2)
library(dplyr)
library(lubridate)
#read in data

rivers <- read_csv('rivers.csv')
wqs <- read_csv('wqs.csv')

#spelling
rivers <- rivers %>%
 mutate( 
    site = case_when(
      site %in% c("Galatin", "Gallatiin", "Gallatin") ~ "Gallatin",
      site %in% c("Madison", "Madison River") ~ "Madison",
      site %in% c("Misouri", "Missouri") ~ "Missouri",
      TRUE ~ site
    ))

#create a df with rivers as variables

rivers_avg <- rivers %>%
   mutate(site = case_when(
    site %in% c("Galatin", "Gallatiin", "Gallatin") ~ "Gallatin",
    site %in% c("Madison", "Madison River") ~ "Madison",
    site %in% c("Misouri", "Missouri") ~ "Missouri",
    TRUE ~ site
  ))%>%
  
  group_by(site) %>%
  summarise(
    avg_temp = mean(temperature, na.rm = TRUE),
    avg_o2   = mean(dissolved_oxygen, na.rm = TRUE),
    avg_ph   = mean(pH, na.rm = TRUE),
    .groups = "drop"
  )

# Display table 

rivers_avg_present <- rivers_avg %>%
  rename(
    "Watershed"        = site,
    "Temperture (C)"= avg_temp,
    "Dissolved O2 (mg/L)"= avg_o2,
    "pH"       = avg_ph
  )



kable(
  rivers_avg_present,
  caption = "Average temperature, Dissolved O2, and pH per watershed",
  digits = 2,          
  format = "markdown"   
)
```

Analysis:

This is a table displaying the average temprture, dissolved oxygen, and the pH for each nearby watershed. All of the watershed average temperatures appear to be within a small temperature range. The Gallatin appears to fall just above that range at 22.35 degrees. This compromises fishery health as you can observe the ammount of disolved oxygen in the gallitan is noticably less than the other rivers. The pH on the other hand apears to be unremarkable.

To get this data, I first fixed spelling errors in the data table. for example, there where three different spellings of "Gallatin" and the maddison had variables like "maddison" and "maddison river" I combined all of those descrepencies into one variable. I then sumarised the data based on the "site" variable and took the mean of each watersheds ph, O2, and temp. I then made a new dataframe with nicer titles and displayed the data using Kable.

```{r}
#| echo: false
# mutate to total the number of times the temp goes above 25 deg

above_25 <- rivers %>%
  mutate(site = case_when(
    site %in% c("Galatin", "Gallatiin", "Gallatin") ~ "Gallatin",
    site %in% c("Madison", "Madison River") ~ "Madison",
    site %in% c("Misouri", "Missouri") ~ "Missouri",
    TRUE ~ site
  ))%>%
  group_by(site) %>%
  summarise(
    above_25 = sum(temperature >= 25, na.rm = TRUE),
    .groups = "drop"
  )

#plot

above_25_plot <- above_25%>%
  ggplot(aes(x = site, y = above_25))+
  geom_col(color = "black", fill = "red")+
  labs(title = "Plot 1: Number of days each watershed averaged above 25 deg (c)",
       x = "Location", 
       y = "Number of days water temp was above 25 (deg C)")
above_25_plot
```

Analysis: This chart is a visualization of the number of days each watershed averaged above 25 deg C, an important exceedence number in fishery health. The Madison is noticeably low, indicating it was able to hold a healthy temperature. Notice how the gallatin is concerning high. This correlates well with the table data.

To get this plot, I ficed the spelling errors, once again and took the sum of when each watershed reached a temperature above or equal to 25 deg C using the sumarize function. I then made a new data frame for the plot and added tiles and colors using ggplot2.

```{r}
#| echo: false
# define date format and create a numeric date column
rivers <- rivers %>%
  mutate(date = parse_date_time(date, orders = c("mdy", "ymd", "dmy")),
         month = month(date)
         )
 rivers_joined <- rivers %>%
  left_join(wqs, by = c("site", "month"))
 
 # total exeeded temps
 
 temp_exceed <- rivers_joined%>%
  group_by(site) %>%
  summarise(
    above_wqs = sum(temperature >= standard, na.rm = TRUE),
    .groups = "drop"
  )
 
 
 #plot
 wqs_plot <- temp_exceed%>%
   ggplot(aes(x = site, y = above_wqs))+
   geom_col(color = "black", fill = "green")+
   labs(title = "Plot 2:Days Each Watershed Exceeded Monthly WQS",
        x = "Watershed",
        y = "Number of Days")
wqs_plot
```

Analysis: The Water Quality Standard (WQS) is a number defined for each river for each mnth that defines a temperature that when exceeded, will compromise river health as dissolved O2 rapidly decreases. This lot is a visulaisation of the number of days each river exceeded the WQS. This offers some suprising data. Notice how the Yellowstone is surprisingly high. I am willing to assume this is because due to the high surface area to depth ratio of the Yellowstone, there is little temperature gradient in the river profile. It stands true that as a fish, the Madison is the place to be.

To get this plot,I first defined the date collumn as a date time collumn so R would know that was the type of data it was working with. I then defined each month numerically and joined the WQS and Rivers data frames into a new data frame with the reference variables "month" and "site." Then I made another data frame summarizing the data and sumung each watershed for the number of days "#temperature \>= standard." I ploted using ggplot2, defined colors and outlines, and added a title.




## Code

```{r}
#| eval: false
#| 
#install and read packages

library(tidyverse)
library(knitr)
library(ggplot2)
library(dplyr)
library(lubridate)
#read in data

rivers <- read_csv('rivers.csv')
wqs <- read_csv('wqs.csv')

#spelling
rivers <- rivers %>%
 mutate( 
    site = case_when(
      site %in% c("Galatin", "Gallatiin", "Gallatin") ~ "Gallatin",
      site %in% c("Madison", "Madison River") ~ "Madison",
      site %in% c("Misouri", "Missouri") ~ "Missouri",
      TRUE ~ site
    ))

#create a df with rivers as variables

rivers_avg <- rivers %>%
   mutate(site = case_when(
    site %in% c("Galatin", "Gallatiin", "Gallatin") ~ "Gallatin",
    site %in% c("Madison", "Madison River") ~ "Madison",
    site %in% c("Misouri", "Missouri") ~ "Missouri",
    TRUE ~ site
  ))%>%
  
  group_by(site) %>%
  summarise(
    avg_temp = mean(temperature, na.rm = TRUE),
    avg_o2   = mean(dissolved_oxygen, na.rm = TRUE),
    avg_ph   = mean(pH, na.rm = TRUE),
    .groups = "drop"
  )

# Display table 

rivers_avg_present <- rivers_avg %>%
  rename(
    "Watershed"        = site,
    "Temperture (C)"= avg_temp,
    "Dissolved O2 (mg/L)"= avg_o2,
    "pH"       = avg_ph
  )


kable(
  rivers_avg_present,
  caption = "Average temperature, Dissolved O2, and pH per watershed",
  digits = 2,          
  format = "markdown"   
)

# mutate to total the number of times the temp goes above 25 deg

above_25 <- rivers %>%
  mutate(site = case_when(
    site %in% c("Galatin", "Gallatiin", "Gallatin") ~ "Gallatin",
    site %in% c("Madison", "Madison River") ~ "Madison",
    site %in% c("Misouri", "Missouri") ~ "Missouri",
    TRUE ~ site
  ))%>%
  group_by(site) %>%
  summarise(
    above_25 = sum(temperature >= 25, na.rm = TRUE),
    .groups = "drop"
  )

#plot

above_25_plot <- above_25%>%
  ggplot(aes(x = site, y = above_25))+
  geom_col(color = "black", fill = "red")+
  labs(title = "bar chart analysis of the number of days each watershed averaged above 25 deg (c)",
       x = "Location", 
       y = "Total number of days wter temp was above 25 (deg C)")
above_25_plot

# define date format and create a numeric date column
rivers <- rivers %>%
  mutate(date = parse_date_time(date, orders = c("mdy", "ymd", "dmy")),
         month = month(date)
         )
 rivers_joined <- rivers %>%
  left_join(wqs, by = c("site", "month"))
 
 # total exeeded temps
 
 temp_exceed <- rivers_joined%>%
  group_by(site) %>%
  summarise(
    above_wqs = sum(temperature >= standard, na.rm = TRUE),
    .groups = "drop"
  )
 
 
 #plot
 wqs_plot <- temp_exceed%>%
   ggplot(aes(x = site, y = above_wqs))+
   geom_col(color = "black", fill = "green")+
   labs(title = "Number of Days Each Watershed Exceeded the Monthly WQS",
        x = "Watershed",
        y = "Number of Days")
wqs_plot
```
